Probability Model

### 2019-06-28 workout_probabilities.Rmd
add latex stuff

---
title: "workout_probabilities"
author: "Brian Mannakee"
date: "4/22/2019"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

# Intro

Variant callers are statistical classifiers.
Usefulness of any classifier is a combination of the ranking performance of the classifier, and the quality of decisions based on the classifier in the real world.
Here we describe an empirical Bayes method which uses very high confidence variant calls from MuTect to create a biologically inspired prior probability of mutation based in the mutation spectrum of the tumor.
We use realistic tumor simulations to show that this method is superior to MuTect as a classifier based on AUROC, while equal or inferior to MuTect based on AUPRC except in situations where the distribution of allele frequencies is extreme and unrealistic.
We then implement a method of false discovery rate control that recovers the superiority of the classifier in the decision context.

@@

We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
Unfortunately, this increased classification performance comes at a cost in terms of decision-making.
As measured by AUPRC, performance is a complex function of sequencing depth, mutation signature concentration, and the allele frequency distribution.
In our simulations there is always a point on the PRC curve where MuTect outperforms our method.
This is a result of the fact that for a given there is an allele frequency at which the ratio of expect true positives to expected false positives from being greater than 1 to less than 1.
There is also an asymmetry in that around any particular threshold, the gradient of the ratio is fairly steep (good lord, this is awful).

# False positive rate control.

We develop a method, following Efron(2008), for controlling the false positive rate.
Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either *null* (non-variant with $\nu = 0$) or *nonnull* (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,

@@

The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
In a neutrally evolving tumor, the count of variants with a given allele frequency is(bozic et al)
$$
N(f) = \frac{N\mu}{f},
$$
Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
This results in the estimated local nonnull probability
$$
\hat{p}_1 = \frac{N(f)}{N} = \frac{\mu}{f}
$$
and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
the rest just copies Efron. Time to implement and try to find out how it performs.

>>>>

---
title: "Probability Model"
author: "Brian Mannakee"
date: "6/25/2019"
output:
    pdf_document:
        fig_caption: yes

bibliography: caller_paper.bib
---

# Introduction

Variant callers are statistical classifiers.
Usefulness of any classifier is a combination of the ranking performance of the classifier, and the quality of decisions based on the classifier in the real world.
Here we describe an empirical Bayes method which uses very high confidence variant calls from MuTect to create a biologically inspired prior probability of mutation based on the mutation spectrum of the tumor.
We use realistic tumor simulations to show that this method is superior to MuTect as a classifier based on AUROC, while equal or inferior to MuTect based on AUPRC.
The model above is poorly calibrated, so we use estimates of the mutation rate and the distribution of allele frequencies under the null hypothesis.
These estimates rely only the well-supported assumption that after initiation the tumor is evolving essentially neutrally.
We show that the model is well-calibrated with this full estimate of the posterior distribution.

@@

We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
Unfortunately, this increased classification performance comes at a cost in terms of calibration, The probabilities from this model are substantially worse than those from Mutect, such that threshold selection is essentially impossible and precision/recall is substantially worse for this model.
There are several distributional assumptions in this model that effect model calibration, and we address all of them below.

# Estimating the mutation rate from the data

As discussed above, MuTect fixes the site-specific mutation probability at $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$.
All variant callers we are aware of either fix this parameter $\mu$ or allow the user to input the value, but there is no way to really know this value until the variant allele frequencies have been observed.
As with estimation of the mutation signature, we can use high confidence mutations and a model of tumor evolution to compute the tumor-specific mutation rate.
@Bozic2016 show that for any variant allele frequency $\alpha$, the total number of mutations with frequency greater than $\alpha$ and less than $0.25$ is

$$
N(\alpha) = N\mu \left( \frac{1}{\alpha} - \frac{1}{0.25} \right)
$$

Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
By selecting $\alpha$ such that we are highly confident in all calls at frequencies greater than $\alpha$, we can compute $\mu$ and recompute the odds in favor of $\mathbf{H_1}$.
This improves the calibration of the model, but it is still worse than MuTect.
The final assumption embedded in the model is that all variants, regardless of allele frequency, have the same prior probability of being true variants.
Below we address this assumption.

# False positive rate control

We develop a method, following @Efron2008, for controlling the false positive rate.
Every site with sufficient coverage and at least 1 alternate read falls into one of two classes, they are either *null* (non-variant with $\nu = 0$) or *nonnull* (variant with $\nu = \hat{f}$) with prior probabilities $p_0$ and $p_1 = 1-p_0$,

@@

The local, or site-specific, true positive probability $p_1$ can be estimated as the fraction of all sequenced sites that are expected to be positive.
In a neutrally evolving tumor, the number of cells is growing exponentially, and the count of variants with an allele frequency greater than a given allele frequency $f$ is [@Bozic2016,@Williams2016a]
$$
N(f) = \frac{N\mu}{f},
$$
Where $N$ is the total number of sites sequenced and $\mu$ is the per-site mutation probability.
The estimated fraction of all of the sites in the genome that will have a mutation with frequency $f$ is
$$
\hat{p}_1 = \frac{\int_{f_{-}}^{f_{+}} N(f)}{N} = \frac{\mu}{f - .1f} -  \frac{\mu}{f + .1f}
$$
and the estimated null probability $\hat{p}_0 = 1 - \hat{p}_1$.
@Williams2016a provides a full derivation for this, we are essentially computing the integral here of $N(f)$ in a small area around $f$.

# Results

Figure 1 shows that the calibration is better than MuTect or the model without false positive rate control.
In addition estimation of mu from the data results in better calibration than assuming a mis-specified rate.

# Figures

![Model performance 500X whole exome for 3 classes of mutation spectrum](figures/wes_left_plot.pdf)

![Allele frequency profile 500X whole exome for 3 classes of mutation spectrum. Frequencies at a threshold p(H1) = 0.8](figures/wes_right_plot.pdf)

![Model performance 100X whole genome for 3 classes of mutation spectrum](figures/wgs_left_plot.pdf)

![Allele frequency profile 100X whole genome for 3 classes of mutation spectrum. Frequencies at a threshold p(H1) = 0.8](figures/wgs_right_plot.pdf)

![Convergence of empirical prior](figures/kl_figure.png)

![Referene input signatures](figures/input_signatures.png)

# References
